---
title: "Utility Knife Blade Consumer Recomendation"
author: "J. C. Thomas"
date: "4/16/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(kableExtra)
library(patchwork)
library(MASS)
library(EMSaov)
library(apa)
library(agricolae)
```

# Research Question

|        Utility blades are a common household tool used in a variety of applications ranging from breaking down cardboard boxes to stripping wire. Rust is a major concern when working with tools. Depending on the implement, rust can make a tool less effective or unusable. In addition to rust, another factor that may contribute to the performance of a utility blade is the brand as specifications for product manufacturing differ. The price of a utility blade may also reflect the quality of the tool. This experiment seeks to discover the relationship between the performance of utility blades and rust, brand, and price. Performance is measured by primary function of the tool, the ability to cut. The goal of this analysis is to inform the consumer on which utility blades the best cutting power.

# Design

|        To answer this question, utility blades are fit into a weighted jig and dragged across a length of 550 paracord. The number of passes it takes to completely sever the cord is the response. As it is more desirable to have a blade that cuts efficiently, a low number of passes to cut the cord is a desired outcome

|        A pack of utility blades is bought from 10 well known brands. Price is broken up into 2 groups (low and high price) based on price per blade. The two price groups contain 5 brands each, representing the 5 lowest and 5 highest priced blades. The brand, the number of blades per pack, pack price, price per blade, and price group are shown in Table 1. 

```{r, echo=FALSE}
blade_price <- read.csv("blade prices (stat5201).csv")
blade_price$group <- c(rep("L", 5), rep("H", 5))
colnames(blade_price) <- c("Brand", "Num. Blades", "Price", "Price Per Blade", 
                           "Price group")
blade_price <- arrange(blade_price, Brand)
kable(blade_price, caption = "Utility Blade Brand and Prices", booktabs = T) |> row_spec(0,bold=TRUE)
```

|        Four blades are randomly selected from each of the packs (each pack being a different brand). The first two blades selected receive the experimental treatment (rust) and the second 2 blades receive the control treatment (no rust). Thus, there are 2 experimental units for each combination of brand and treatment. Table 2 shows the results of the random treatment assignment where the value $i$ represents the $i^{th}$ blade from the front of the pack. This $i^{th}$ blade is selected for a given treatment. The details of the randomization procedure can be found in the code appendix.

```{r, echo=FALSE}
set.seed(826)
dewalt <- sample(1:10, size = 4, replace = FALSE)
dewalt_c <- sample(1:5, size = 4, replace = FALSE)
hart <- sample(1:5, size = 4, replace = FALSE)
husky <- sample(1:10, size = 4, replace = FALSE)
hyper_tough <- sample(1:5, size = 4, replace = FALSE)
irwin <- sample(1:5, size = 4, replace = FALSE)
kobalt <- sample(1:5, size = 4, replace = FALSE)
lenox <- sample(1:5, size = 4, replace = FALSE)
milwaukee <- sample(1:5, size = 4, replace = FALSE)
stanley<- sample(1:5, size = 4, replace = FALSE)
rand_groups <- rbind("DeWalt" = dewalt, "Dewalt Carbide" = dewalt_c,
                     "Hart" = hart, "Husky" = husky, "Hyper Tough" = hyper_tough, 
                     "Irwin" = irwin, "Kobalt" = kobalt, "Lenox" = lenox, 
                     "Milwaukee" = milwaukee, "Stanley" = stanley) |> as.data.frame()
colnames(rand_groups) <- c("Rust", "Rust", "Control", "Control")
kable(rand_groups, caption = "Treatment Assignment", booktabs = T) |> row_spec(0,bold=TRUE)
```

|        The two utility blades selected for the rust treatment are sprayed with a rusting solution of 150ml of vinegar, 150ml of hydrogen peroxide, and 5 teaspoons of salt. The blades are left outside on a piece of cardboard for 8 hours. They are flipped and resprayed with the solution every two hours. After the treatment, the blades are brought inside, dried, and set on a new piece of cardboard. They sit for another 12 hours before the testing begins. 

|        The four of blades for each brand were assigned a blade number 1 though 4 with 1 and 2 corresponding rusted blades and 3 and 4 corresponding to control blades. All 40 blade number/ brand combinations were randomly placed into a testing order. The results of this randomization are found in Table 3 where the column "Utility Blade" contains the name of a brand followed by a blade number. The column "Order" contains the order in which the blades are tested. Details of the randomization process can be found in the code appendix.

```{r, echo=FALSE}
set.seed(826)
group_paste <- matrix(rep(NA, 40), nrow = 10, ncol = 4)
for (i in 1:nrow(group_paste)) {
  group_paste[i,] <- paste(rownames(rand_groups)[i], 1:4, "")
}
group_vec <- c()
for (i in 1:nrow(group_paste)) {
  group_vec <- c(group_vec, group_paste[i, ])
}
blade_order <- sample(group_vec, 40)
order_table <- data.frame("Order" = 1:40,
                          "Utility Blade" = blade_order)
#write.csv(order_table, "Xblade order and collection (stat5101).csv")
colnames(order_table) <- c("Order", "Utility Blade")
kable(list(order_table[1:20,], order_table[21:40,]), 
      caption = "Testing Order", booktabs = T)
```

|        Each blade is individually secured into a jig to remove operator variation between passes of the blade. The jig consists of half inch PVC pipe secured to wheels. A 5lb chain weights down the front of the jig where the utility blade is held to ensure sufficient pressure is applied. A piece of 550 paracord is stretched over a piece of melamine, pulled tot, and stapled twice on each side to secure it to the melamine. The blade passes over the 1.5 inch tot section of cord secured between the sets of staples. Each blade was tested only one time to prevent any variation in the response from dulling.

|        The factors of this experiment are treatment (rust or no rust), brand (see 10 brands in Table 1), and price group (high or low), all of which are fixed factors. Brand/treatment and price group/treatment are both crossed and brand is nested within price group. Each brand and treatment combination is replicated twice.

# Data

```{r, echo=FALSE, message=FALSE}
blade <- read.csv("blade order and collection (stat5201).csv")
blade <- blade[-1]
colnames(blade)[3] <- "pass"
blade <- arrange(blade, Utility.Blade)
blade$brand <- str_extract(blade$Utility.Blade, "^[A-Za-z]+ *[A-Za-z]*") |>
  trimws()
blade$trt <- rep(c("Rust", "Rust", "Control", "Control"), 10)
blade <- left_join(blade, blade_price, by = c("brand" = "Brand"))
blade_res <- blade[-c(1,2,6:8)]
blade_res <- blade_res |> group_by(brand, trt, `Price group`) |>
  summarise("con" = paste(pass, collapse = ", ")) |> 
  pivot_wider(names_from = trt, values_from = con)
colnames(blade_res)[c(1,2)] <- c("Brand", "Price Group")
res_tab <- kable(blade_res[c(2,1,3,4)], booktabs = T, caption = "Results") |>
  add_header_above(c(" "=2, "Treatment" = 2)) |>
  kable_styling(position = "center") |> row_spec(0,bold=TRUE)
```

|        The data collected from this experiment can be found in the appendix of this document and is presented graphically in this section. Diagram 1 shows the results by brand where the color designates the treatment applied to the utility blade. All brands seem to be centered around the 5 pass mark besides Hart and Stanley. These two brands require more passes to cut the cord and have a higher variance of passes. The distribution of the results is visualized in Diagrams 2 through 4. Diagrams 2 and 3 show that both price groups and both treatment groups have similar medians. The variance for the low price group is higher than that of the high price group. The variance of the control treatment is higher than that of the rust treatment. Diagram 4 shows the distribution of the overall number of passes required to cut the cord. The mass of this distribution fall beneath 10 passes and the distribution is right skewed.

```{r, echo=FALSE, fig1, fig.height = 3, fig.width = 10, fig.align = "center"}
colnames(blade) <- c("order", "util_blade", "pass", "brand", "trt", 
                     "num_blade", "price", "ppb", "p_group")
#
p0 <- ggplot(blade) + 
  geom_point(aes(x = brand, y = pass, col = trt), 
             position = position_jitter(width = .05)) +
  labs(title = "Diagram 1: Results by Brand")
p0
```

```{r, echo=FALSE, fig2, fig.height = 4, fig.width = 7, fig.align = "center"}
#
p1 <- ggplot(blade) + 
  geom_boxplot(aes(x = p_group, y = pass), outlier.color = NA)
p1 <- p1 + geom_point(aes(x = p_group, y = pass, col = p_group),
                position = position_jitter(width = .05), alpha = .5)+
  labs(title = "Diagram 2: Distribuiton Between Price Groups",
       x = "Price Group", y = "# Passes")+ 
  theme(plot.title = element_text(size = 10), legend.position="none")
#
p2 <- ggplot(blade) + geom_boxplot(aes(x = trt, y = pass), outlier.color = NA)
p2 <- p2 + geom_point(aes(x = trt, y = pass, col = trt),
                      position = position_jitter(width = .05), alpha = .5) +
  labs(title = "Diagram 3: Distribuition Between Treatments",
       x="Treatment", y = "# Passes")+ 
  theme(plot.title = element_text(size = 10), legend.position="none")
#
p3 <- ggplot(blade, aes(x = pass)) +
   geom_histogram() +
  labs(title = "Diagram 4: Distribution of All Results",
       x = "# Passes", y = "Count") + 
  theme(plot.title = element_text(size = 10))
((p1 + p2) / p3 + plot_layout(heights = c(2,1))) |> print() |> suppressMessages()
```

# Analysis

|        The structure of this experiment consists of three fixed factors: treatment, price group, and brand. Brand is nested within price group and will be notated brand[price group]. Treatment and price group are crossed and so is treatment and brand[price group]. The model used to fit this structure is as follows:

$$y_{ijkm} = \mu + \alpha_i + \beta_j + \gamma_{k[j]} + (\alpha \beta)_{i,j} + (\alpha\gamma)_{i,k[j]} + \epsilon_{m[i,k[j]]}$$
with

 * $\mu$: grand mean
 * $\alpha_i$: $i^{th}$ level of treatment
 * $\beta_j$: $j^{th}$ level of price group
 * $\gamma_{k[j]}$: $k^{th}$ level of brand[price group]
 * $\epsilon_{m[i,k[j]]}$: error.
 
|        The full model is fit to the data under the typical assumptions of error independence, homoscedasticity, and error normality. Diagnostic plots for this fit are given in Diagram 5. The glaring assumption violation in Diagram 5 is the non-constant variance on the plot "Residuals vs Fitted". This indicates a violation of homoscedasticity. The Box-Cox procedure (Diagram 6) applied to the full model, shows that a negative square root power transformation may help stabilize the variance. This transformation is performed on the response variable and the full model is fit on the transformed responses. The diagnostic plots for the full model on the transformed responses are shown in Diagram 7. The diagnostic plots in Diagram 7 indicate constant variance (no patterns in "Residuals vs Fitted") and normality of errors (adherence to the normal line in "Normal Q-Q") assumptions are both met. Checking for outliers by comparing the absolute value of the externally studentized residuals to a Bonferroni corrected t-value indicates there are no outliers.

```{r, echo=FALSE,fig3, fig.height = 4, fig.width = 4, fig.align = "center"}
#fit full model
blade_m0 <- aov(pass ~ trt + p_group + brand%in%p_group + trt:p_group +
                  trt:(brand%in%p_group), data = blade)
par(mfrow=c(2,2))
plot(blade_m0) |> suppressMessages()
```

```{r, echo=FALSE, fig4, fig.height = 3, fig.width = 3, fig.align = "center"}
#boxcox
par(mfrow=c(1,1))
boxcox(blade_m0)
```
```{r, echo=FALSE, fig5, fig.height = 4, fig.width = 4, fig.align = "center"}
#transform response and fit full model
blade$t_pass <- (blade$pass)^(-.5)
blade_m1 <- aov(t_pass ~ trt + p_group + brand%in%p_group + trt:p_group +
                  trt:(brand%in%p_group), data = blade)
par(mfrow=c(2,2))
plot(blade_m1) |> suppressMessages()
#checking for outliers
ex_resids_m1 <- rstudent(blade_m1)
t_resid <- qt(.05/(2*nrow(blade)), df = 8) #df here df of highest interaction
num_res <- sum(t_resid > abs(ex_resids_m1), na.rm = TRUE)
```


|        Interaction plots for the treatment/price group interaction and treatment/brand[price group] interaction are shown in Diagram 8 and 9 respectively. Diagram 8 shows no evidence of interaction between treatment and price group. There are some intersecting lines in Diagram 9 indicating that some of the treatment/brand[price group] interactions may be significant. Due to limitations with the software, not all brand names could be shown on the x axis however they are in the same order as in Table 1. The control/DeWalt Carbide, control/Hart, and control/Stanley interactions show overlap or sudden trend variation thus indicating possible 1 cell interaction.

```{r, echo=FALSE, fig6, fig.height = 4, fig.width = 8, fig.align = "center"}
#interaction plots
interaction.plot(x.factor = blade$trt,
                 trace.factor = blade$p_group, 
                 response = blade$t_pass,
                 ylab = "# Passes (transformed",
                 xlab = "Treatment",
                 trace.label = "Price Group",
                 main = "Diagram 8: Interaction Plot for Price Group and Treatment")
```
```{r, echo=FALSE, fig7, fig.height = 4, fig.width = 8, fig.align = "center"}
interaction.plot(x.factor = blade$brand,
                 trace.factor = blade$trt, 
                 response = blade$t_pass,
                 ylab = "# Passes (transformed",
                 xlab = "Brand",
                 trace.label = "Treatment",
                 main = "Diagram 9: Interaction Plot for Treatment and Brand")
```

|        An ANOVA is ran for the full model on the transformed data yielding the results in Table 4. As all the effects in this model are fixed, the leading eligible random term is the error. This means that all tests will be performed with the MSE as the denominator. Table 4 indicates the interaction of treatment/brand[price group] is not significant, further giving support to possible one cell interactions. The one cell interactions previously specified are modeled as dummy variables and added to the transformed full model. They are each added individually to avoid problems with non-orthogonal columns. All of the dummy variables representing one cell interactions are significant when placed alone in the transformed full model. Thus, they are added in together to the transformed full model yielding the ANOVA table in Table 5. Note that "Dum. C/DC", "Dum. C/H", and "Dum. C/S" represent the dummy variables used for modeling the one cell interaction in control/Dewalt Carbide, control/Hart, and control/Stanley respectively.

```{r, echo=FALSE}
#transformed anova
sm1 <- summary(blade_m1)
df_sm1 <- matrix(unlist(sm1), ncol = 5) |> round(4) |>
  as.data.frame()
colnames(df_sm1) <- c("DF", "SS", "MS", "F stat", "P-val")
rownames(df_sm1) <- c("Treatment", "Price Group", 
                       "Brand[Price Group]",
                      "Treatment * Price Group", 
                      "Treatment * Brand[Price Group]",
                      "Residuals")
df_sm1$Sig. <- c("", "***","***","", "", "")
df_sm1[6,4] <- ""
df_sm1[6,5] <- ""
kable(df_sm1, caption = "Transformed Full Model ANOVA", booktabs = T) |> row_spec(0,bold=TRUE)
```

```{r, include=FALSE}
#create dummy variable for control/hart
blade$d_ch <- rep(0, nrow(blade))
blade$d_ch[c(11,12)] <- 1
blade_m2 <- aov(t_pass ~ trt + p_group + brand%in%p_group + trt:p_group +
                  trt:(brand%in%p_group) + d_ch, data = blade)
summary(blade_m2)
#create dummy variable for control/stanley
blade$d_cs <- rep(0, nrow(blade))
blade$d_cs[c(39, 40)] <- 1
blade_m3 <- aov(t_pass ~ trt + p_group + brand%in%p_group + trt:p_group +
                  trt:(brand%in%p_group) + d_cs, data = blade)
summary(blade_m3)
#create dummy variable for control/dewalt carbide
blade$d_cdc <- rep(0, nrow(blade))
blade$d_cdc[c(7,8)] <- 1
blade_m4 <- aov(t_pass ~ trt + p_group + brand%in%p_group + trt:p_group +
                  trt:(brand%in%p_group) + d_cdc, data = blade)
summary(blade_m4)
```

```{r, echo=FALSE}
#use all dummies together
blade_m5 <- aov(t_pass ~ trt + p_group + brand%in%p_group + trt:p_group +
                  trt:(brand%in%p_group) + d_ch + d_cs + d_cdc, data = blade)
sm5 <- summary(blade_m5)
df_sm5 <- matrix(unlist(sm5), ncol = 5) |> round(4) |>
  as.data.frame()
colnames(df_sm5) <- c("DF", "SS", "MS", "F stat", "P-val")
rownames(df_sm5) <- c("Treatment", "Price Group", "Dum. C/H",
                      "Dum. C/S", "Dum. C/DC", "Brand[Price Group]",
                      "Treatment * Price Group", "Treatment * Brand[Price Group]",
                      "Residuals")
df_sm5$Sig. <- c("", "***", "**", "**", "*", "*", "", "", "")
df_sm5[9,4] <- ""
df_sm5[9,5] <- ""
kable(df_sm5, caption = "Transformed Full Model with Dummy Variables ANOVA",
      booktabs = T) |> row_spec(0,bold=TRUE)
```

|        The p-value for the treatment/brand[price group] interaction drastically decreases (from 0.87 to 0.99)  when the dummmy variables are included. This means that accounting for the 3 once cell interactions eleminates almost all of the interaction effect from the treatment/brand[price group] interaction. This makes removing the full interaction term much more palatable. 

|        With interaction concerns nullified, backwards selection based on F value is conducted to find the most parsimonious model for the data. After only one round of elimination, the best model is found to include the three main effects (treatment, price group, and brand[price group]) and the effects of the dummy variables modeling the one cell interactions. All of these effects are significant at the $\alpha = 0.05$ level besides the effect of treatment which is significant only at the $\alpha = 0.1$ level. This indicates that the treatment effect is not significant. Adding to its non-significance is the fact that multiple tests are being performed in the ANOVA simultaneously. This falsely inflates the F statistic. However, in order for the model to remain hierarchical, the treatment effect must remain as it is used in the dummy variables modeling the one cell interactions. The ANOVA for this model is found in Table 6.

```{r, include=FALSE}
#drop treatment/brand
blade_m6 <- aov(t_pass ~ trt + p_group + brand%in%p_group + trt:p_group +
                  d_ch + d_cs + d_cdc, data = blade)
summary(blade_m6)
#drop trt/price group
blade_m7 <- aov(t_pass ~ trt + p_group + brand%in%p_group +
                  d_ch + d_cs + d_cdc, data = blade)
summary(blade_m7)
```

```{r, echo=FALSE}
sm7 <- summary(blade_m7)
df_sm7 <- matrix(unlist(sm7), ncol = 5) |> round(4) |>
  as.data.frame()
colnames(df_sm7) <- c("DF", "SS", "MS", "F stat", "P-val")
rownames(df_sm7) <- c("Treatment", "Price Group", "Dum. C/H",
                      "Dum. C/S", "Dum. C/DC", "Brand[Price Group]",
                      "Residuals")
df_sm7$Sig. <- c(".", "***", "***", "***", "**", "**", "")
df_sm7[7,4] <- ""
df_sm7[7,5] <- ""
kable(df_sm7, caption = "Final Model ANOVA", booktabs = T) |> row_spec(0,bold=TRUE)
```

|        With significant terms identified, post hoc investigation into the effects is performed to identify which levels of each factor differ significantly. This is be done through the use of pairwise contrasts. Contrast significance is evaluated with the Tukey HSD test at level $\alpha = 0.05$. Table 6 displays the brands that differ significantly from each other alongside the transformed mean and standard deviation for each brand. Any two brands not connected by a column of 1s are significantly different from one another. An example of this is the Stanley and Hart brands being connected by a column of 1s and therefore not being significantly different. However, Hart and Lenox are not connected by a column of 1s indicating they differ significantly. Pairwise contrasts are not necessary to determine which groups are significantly different in the price group effect as there are only two levels. By looking at the averages and the high P-value in Table 6 it is evident that the mean of the high price group (0.506) is significantly different than the mean of the low price group (0.397). 

```{r, include=FALSE}
#price group means
blade |> group_by(p_group) |> summarise("avg" = mean(t_pass))
```


```{r, echo=FALSE}
#contrast for  brand
brand_con <- HSD.test(blade_m7, "brand")$groups
brand_con$A <- c(rep(1, 7), 0,0,0)
brand_con$B <- c(0, rep(1, 7), 0, 0)
brand_con$C <- c(rep(0, 7), 1, 1, 0)
brand_con$D <- c(rep(0, 8), 1, 1)
colnames(brand_con)[1] <- "Avg. # Passes (transformed)"
brand_con <- brand_con[-2]
brand_con$St.D. <- HSD.test(blade_m7, "brand")$means[c(8, 1, 7, 9, 5, 4, 6, 2, 10, 3), 2] |>
  round(3)
brand_con$`Avg. # Passes (transformed)` <- round(brand_con$`Avg. # Passes (transformed)`, 3)
kable(brand_con[c(1,6,2,3,4,5)], caption = "Brand Contrasts", booktabs = T) |> row_spec(0,bold=TRUE)
```

# Conclusion

The final model given in the analysis section sheds light onto which utility blades are best for a consumer to buy. The factors considered in this experiment were rust, price range, and brand. It was found that whether or not a blade is rusty does not have a significant effect on a blade's ability to cut. Thus, marketing claims about a particular blade being resistant to rust need not be a  principle factor when selecting a utility blade. There are only 3 brands that saw effect from rusting. This effect was a decrease in cutting efficiency meaning that if a blade is being purchased for use in primarily wet environments (ie work on pools or boats), the Hart, DeWalt Carbide, and Stanley brands should be avoided. Beyond that, the brand purchased for work in wet applications does not matter.

The brand itself, however, is an important factor. When making these comparisons, it is important to recall that this analysis was carried out on the responses with a negative square root power transformation. This means that large values of the transformed response are desirable over small values. The Lenox brand came out on top with an average of 0.610 transformed passes to cut the cord (2.75 untransformed). The Lenox brand is the second most expensive brand tested making it not a budget friendly option. Via pairwise comparisons, it was determined that the Lenox brand has no discernible difference in cutting performance from the Husky brand, which is second least expensive. The Husky brand is a more economical brand that will yield the same level of performance. The two brands that performed the worst were Stanley and Hart. Hart was the clear loser in this match up averaging 0.257 transformed passes (17 untransformed). Although having a slightly better average, Stanley utility blades do not offer significantly superior performance. This is expected from a considerably new brand like Hart but disappointing from Stanley as it is a well established company.

The most significant factor in utility blade purchasing decisions should be the price range. With this product, the old adage of "you get what you pay for" rings true. Blades from the higher price range perform better. This shows in the average transformed passes until severing the cord. The high price group averaged 0.506 transformed passes and the low price group averaged 0.397 transformed passes (4.15 and 8.9 untransformed, respectively). There is no interaction effect between the price group factor and the rust factor meaning there is no advantage in purchasing a more or less expensive blade in hopes of better performance in wet conditions.

Further work could be done on this subject using dulling blades as a factor. How well a blade is able to keep its edge is an important characteristic. Another consideration for a future experiment might be a more corrosive rusting agent. While the solution used in this experiment worked quickly, it is possible that stronger solutions leading to more rust damage could hamper blade performance. Should this experiment be repeated, an increase in sample size for each brand would be beneficial. Perhaps focusing on a few brands with large sample size would be a good path to pursue.

According to this analysis, the highest performing blades are associated with higher prices and specific brand names. Although not purchased too often by most consumers, it is helpful to have a effective utility blade in a knife when it is needed. Additionally, contractors who purchase many utility blades for their teams may benefit from the results of this data. If contractors have previously been purchasing a high priced, high performance blades like the Lenox, it would save them money to purchase blades from a brand like Husky that has equivalent performance at a lower cost.